---
title: "Bioinformatics Project"
author: "Sergio Fernandez - Alessandro Mecchia - Paolo Nicolet"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_depth: 2
---

# Imports and packages

```{r}
Sys.which("pandoc")

BiocManager::install("recount3")

```

# Project Introduction

**da completare**

# Dataset Selection

Load recount3 library and the list of its projects to explore the projects.

recount3 is a large-scale resource of uniformly processed RNA-seq data for humans and mice. It aggregates publicly available datasets and provides them in a standardized format for downstream analysis. **Recount3 contain:**

-   **Sequence Read Archive (SRA)**
    -   Largest source of raw RNA-seq data.
    -   Includes diverse studies across tissues, conditions, and species.
    -   Data are annotation-agnostic and processed uniformly using the Monorail pipeline.
-   **Genotype-Tissue Expression (GTEx) Project**
    -   High-quality RNA-seq data from multiple human tissues.
    -   Focused on understanding tissue-specific gene expression.
-   **The Cancer Genome Atlas (TCGA)**
    -   RNA-seq data from various cancer types.
    -   Enables cancer-specific expression and splicing analyses.

```{r}
library(recount3)

# List of available projects
projects <- available_projects()
head(projects)

```

```{r}
# Filter by GTEx (normal tissues)
gtex_projects <- subset(projects, file_source == "gtex")

# Explore metadata
head(gtex_projects)
```
